---
title: "Outlier_Removal"
author: "Ainhoa Calleja Rodriguez"
date: "2025-01-27"
output: 
 html_document:
   self_contained: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
# Load the YAML configuration file
working_path <- "~/Documents/INIA/analysis_ainhoa/analysis_yaml/"
config <- yaml::read_yaml(paste0(working_path, "P21GUD1.yaml"))

```

## Outlier removal for `r config$trial_name`:
Based on preview descriptive statistics of raw data, `r config$outlier_groups$IQR_vars` are moderately skewed. The remaining variables (`r config$outlier_groups$SD_vars`) are normally distributed or slightly skewed (either right or left). Thus, outlier removal for moderately skewed variables was performed by IQR outlier removal, whereas for the remaining variables SD outlier (`r config$outlier_k`*SD) removal was performed.

```{r, echo=FALSE, message=FALSE, warning=FALSE, results = 'hide'}

# Load functions needed. bafun.R is a script with generic functions to create basic statistics and graphs
bafun_path <- "~/Documents/INIA/analysis_ainhoa/bafun.R"   
source(bafun_path)

# Load (install) packages
packages <- c("yaml","ggplot2", "dplyr", "readxl","tidyr", "viridis", "tidyverse", "summarytools", "gridExtra", "grid", "scales")
lapply(packages, install_and_load)


trial_name <- config$trial_name

# Check the file extension
input_path <- config$input_path
# sheet_name <- config$sheet_name
if (grepl("\\.(xls|xlsx)$", input_path, ignore.case = TRUE)) {
  # If the file is an Excel file
  sheet_name <- config$sheet_name
  df <- read_excel(input_path, sheet = sheet_name, na = c(".Z", ".Q", ".z", ".q"))
} else if (grepl("\\.csv$", input_path, ignore.case = TRUE)) {
  # If the file is a CSV file
  df <- read.csv(input_path, stringsAsFactors = TRUE)
} else {
  stop("Unsupported file format. Please provide a .xlsx or .csv file.")
}

# Modify the data types of selected columns based on YAML configuration
integer_columns <- config$column_transformations$integer_columns
numeric_columns <- setdiff(names(df), c(integer_columns, config$column_transformations$numeric_columns$exclude))

df <- df %>%
  mutate(
    across(all_of(integer_columns), as.integer),
    across(all_of(numeric_columns), as.numeric)
  )

str(df)

```

### IQR outlier detection and removal. 
**1.5 × IQR** as standard statistical rule for this dataset. Tukey's rule: outliers are values more than 1.5 times the interquartile range from the quartiles — either below Q1 − 1.5IQR, or above Q3 + 1.5IQR. 
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
# IQR outliers
IQR_outliers <- function(x) {
  if (is.numeric(x)) {
    Q1 <- quantile(x, 0.25, na.rm = TRUE)
    Q3 <- quantile(x, 0.75, na.rm = TRUE)
    IQR <- Q3 - Q1
    # Set values outside 1.5 * IQR to NA
    x[x < (Q1 - 1.5 * IQR) | x > (Q3 + 1.5 * IQR)] <- NA
  }
  return(x)
}

df[config$outlier_groups$IQR_vars] <- lapply(df[config$outlier_groups$IQR_vars], IQR_outliers)


```

### SD outlier detection and removal 
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}

# SD outliers
SD_outliers <- function(x) {
  if (is.numeric(x)) {
    mean_x <- mean(x, na.rm = TRUE)
    sd_x <- sd(x, na.rm = TRUE)
    x[abs(x - mean_x) > config$outlier_k * sd_x] <- NA
  }
  return(x)
}

# Apply the function SD_outliers
df[config$outlier_groups$SD_vars] <- lapply(df[config$outlier_groups$SD_vars], SD_outliers)

write.csv(df, file = paste0(config$output_path, config$output_file), row.names = FALSE)

```

## Height measurements
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
H_vars <- grep(config$variable_groups$H_vars, colnames(df), value = TRUE)
desired_order <- config$desired_order
H_vars <- desired_order[desired_order %in% H_vars]
descr(df[, H_vars])
H_summ_stats <- calculate_summary(df, group_var = "PROV", target_vars = H_vars)

```

### Boxplots of height measured in different years
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=6, fig.align='center'}
H_graphs <- create_graphs(df = df, 
                        group_var = "PROV", 
                        target_vars = H_vars, 
                        var_label = "Height", 
                        unit_label = "cm", 
                        ncol_p2 = 4,  # 4 columns for p2
                        ncol_p3 = 5,  # 2 columns for p3
                        ncol_p4 = 3,  # 5 columns for p4
                        ncol_p5 = 3)  # 3 columns for p5

H_graphs$Boxplot_Variable

```

### Boxplots of height-year and provenance
```{r,  echo=FALSE, message=FALSE, warning=FALSE, fig.width=15, fig.height=13, fig.align='center'}
H_graphs$Boxplot_Provenance1

```

### Boxplots height-year, per provenance
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=15, fig.height=13, fig.align='center'}
H_graphs$Boxplot_Provenance2
```

### Histograms height-year 
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=11, fig.height=17, fig.align='center'}
# Histogram
H_graphs$Histogram

```

### Density plots of height-year
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=11, fig.height=17, fig.align='center'}
H_graphs$Density

```

## Base diameter (DB) measurements
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
DB_vars <- grep(config$variable_groups$DB_vars, colnames(df), value = TRUE)
DB_summ_stats <- calculate_summary(df, group_var = "PROV", target_vars = DB_vars)
descr(df[, DB_vars])

```

### Boxplots of DB measured in different years
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=3, fig.height=4, fig.align='center'}
# Create boxplot
DB_graphs <- create_graphs(df = df,
                        group_var = "PROV",
                        target_vars = DB_vars,
                        var_label = "DB",
                        unit_label = "mm",
                        ncol_p2 = 4,
                        ncol_p3 = 5,
                        ncol_p4 = 2,
                        ncol_p5 = 2)
DB_graphs$Boxplot_Variable

```

### Boxplots of DB-year and provenance
```{r,  echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=5, fig.align='center'}
DB_graphs$Boxplot_Provenance1

```

### Boxplots DB-year, per provenance
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=9, fig.align='center'}
DB_graphs$Boxplot_Provenance2
```

### Histograms of DB-year
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3, fig.align='center'}
# Histogram
DB_graphs$Histogram

```

### Density plots of DB-year
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3, fig.align='center'}
DB_graphs$Density

```

## Diameter at Breast Height (DBH) and DN measurements
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}

DBH_vars <- grep(config$variable_groups$DBH_vars, names(df), value = TRUE)
# DN_vars <- grep("^DN", names(df), value = TRUE)  #all variables starting with DB
#summary(df[,DBH_vars])
# summary(df[,DN_vars])
descr(df[, DBH_vars])
# descr(df[,DN_vars])
DBH_summ_stats <- calculate_summary(df, group_var = "PROV", target_vars = DBH_vars)  
#head(H_summ_stats)

```

### Boxplots of DBH and DN measured in different years
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=4, fig.align='center'}
# Create boxplot
#DBH_graphs <- create_graphs(df, group_var = "PROV", target_vars = DBH_vars, var_label = "Diameter at Breast Height (DBH)", unit_label = "mm")
DBH_graphs <- create_graphs(df = df, 
                        group_var = "PROV", 
                        target_vars = DBH_vars, 
                        var_label = "Diameter at Breast Height (DBH)", 
                        unit_label = "mm", 
                        ncol_p2 = 4,  # 4 columns for p2
                        ncol_p3 = 5,  # 2 columns for p3
                        ncol_p4 = 4,  # 5 columns for p4
                        ncol_p5 = 4)  # 3 columns for p5
DBH_graphs$Boxplot_Variable

# DN_graphs <- create_graphs(df = df,
#                            group_var = "PROV", 
#                            target_vars = DN_vars, 
#                            var_label = "Diameter Forked Trees (DN)", 
#                            unit_label = "mm",
#                            ncol_p2 = 4,
#                            ncol_p3 = 5,
#                            ncol_p4 = 1,
#                            ncol_p5 = 1)
# DN_graphs$Boxplot_Variable

```

### Boxplots of DBH-year and provenance, DN
```{r,  echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=5, fig.align='center'}
DBH_graphs$Boxplot_Provenance1
#DN_graphs$Boxplot_Provenance1
```

### Boxplots DBH-year, per provenance
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=9, fig.align='center'}
DBH_graphs$Boxplot_Provenance2
#DN_graphs$Boxplot_Provenance2
```

### Histograms of DBH-year
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=3, fig.align='center'}
# Histogram
DBH_graphs$Histogram
#DN_graphs$Histogram

```

### Density plots of DBH-year and DN
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=3, fig.align='center'}
DBH_graphs$Density
#DN_graphs$Density
```

## Scatter plots H vs DB
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=18, fig.align='center'}
# Extract DBH and H, and get the last two digits
rel_vars <- grep("^H|^DB", names(df), value = TRUE)
var_digits <- gsub(".*(\\d{2})$", "\\1", rel_vars)
# new df
df_vars <- data.frame(variable = rel_vars, digits = var_digits)
# Group by the digits 
df_grouped <- df_vars %>%
  group_by(digits) %>%
  arrange(variable) %>%
  mutate(pair = list(variable))

# Create the plots
plots <- list()

for(i in 1:nrow(df_grouped)) {
  pair <- df_grouped$pair[[i]]
  
  # Check if both H and DB variables exist before attempting to plot
  var1 <- pair[1]
  var2 <- pair[2]
  
  # If there is no match for the digits of DB or H, use the previous 
  if(length(pair) == 1) {
    var2 <- ifelse(grepl("^H", var1), gsub("H", "DB", var1), gsub("DB", "H", var1))
    if (!var2 %in% names(df)) next  # Skip if the second variable doesn't exist
  }
  
  # Paired plots
  if (var1 %in% names(df) && var2 %in% names(df)) {
    p <- ggplot(df, aes_string(x = var1, y = var2)) + 
      geom_point(size = 1) + 
      geom_smooth(method = "lm") + 
      ggtitle(paste(trial_name)) + 
      xlab(paste(var1)) + 
      ylab(paste(var2)) + 
      scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) + 
      scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    
    plots[[length(plots) + 1]] <- p
  }
}


if (length(plots) > 0) {
  grid.arrange(grobs = plots, ncol = 2)
} else {
  print("No valid plots to display.")
}

```

## Paired scatterplots H
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=22, fig.align='center'}
paired_scatter(df, H_vars, ncol = 3)
```

## Paired scatterplots DBH
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=3, fig.align='center'}

paired_scatter(df, DBH_vars, ncol = 3)


```
