---
title: "RRegression_singleSite"
author: "Ainhoa Calleja Rodriguez"
date: "2025-02-07"
output:
  html_document:
    self_contained: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
# Load the YAML configuration file
working_path <- "~/Documents/INIA/analysis_ainhoa/analysis_yaml/"
config <- yaml::read_yaml(paste0(working_path, "P21NAV.yaml"))
```

# Single Site Random Regression Analysis for `r config$trial_name` (script still under development).

The variables analised in this script were `r config$desired_order`. This variables represent repeated measurements of the same variable across multiple years (i.e., same individuals recorded multiple times). Thus random effect regression was chosen as the method to analysed these repeated measurements. The goal was to obtain a linear regression for each provenance across years, and to obtain the temporal correlation. 


```{r, echo=FALSE, message=FALSE, warning=FALSE, results = 'hide'}

#Load functions
bafun_path <- "~/Documents/INIA/analysis_ainhoa/bafun.R"
source(bafun_path)

# Load (install) necessary packages
packages <- c("ggplot2", "dplyr", "readxl","tidyr", "viridis", "tidyverse", "gridExtra", "grid", "scales", "asreml", "ggrepel", "kableExtra")
lapply(packages, install_and_load)

# Load data according to file extension
trial_name <- config$trial_name
input_path <- config$input_path
sheet_name <- config$sheet_name

if (grepl("\\.(xls|xlsx)$", input_path, ignore.case = TRUE)) {
  # Excel file
  sheet_name <- config$sheet_name
  df <- read_excel(input_path, sheet = sheet_name, na = c(".Z", ".Q", ".z", ".q"))
} else if (grepl("\\.csv$", input_path, ignore.case = TRUE)) {
  # CSV file
  df <- read.csv(input_path, stringsAsFactors = TRUE)
} else {
  stop("Unsupported file format. Please provide a .xlsx or .csv file.")
}
str(df)
names(df)

# Columns to exclude
exclude_cols <- config$exclude_columns
df <- df[, !(names(df) %in% exclude_cols)]
names(df)

# column renaming
if (!is.null(config$rename_columns$specific_renames)) {
  for (old_name in names(config$rename_columns$specific_renames)) {
    new_name <- config$rename_columns$specific_renames[[old_name]]
    if (old_name %in% colnames(df)) {
      colnames(df)[colnames(df) == old_name] <- new_name
    }
  }
}

# Pattern renaming
if (!is.null(config$rename_columns$pattern_replacements)) {
  for (pattern in names(config$rename_columns$pattern_replacements)) {
    replacement <- config$rename_columns$pattern_replacements[[pattern]]
    colnames(df) <- gsub(pattern, replacement, colnames(df))
  }
}

names(df)

# column type change based on YAML
integer_columns <- config$column_transformations$integer_columns
numeric_columns <- setdiff(names(df), c(integer_columns, config$column_transformations$numeric_columns$exclude))

df <- df %>%
  mutate(
    across(all_of(integer_columns), as.integer),
    across(all_of(numeric_columns), as.numeric)
  )

str(df)

```

Check missing values:

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
# NA values for columns with lower H values than previous years
height_cols <- grep("^H\\d{2}$", names(df), value = TRUE)
na_count <- colSums(is.na(df[ , height_cols]))
non_na_count <- colSums(!is.na(df[ , height_cols]))
check_NAs <- data.frame(Column = height_cols, NA_Count = na_count, Non_NA_Count = non_na_count)
check_NAs
# for (i in 2:length(height_cols)) {
#   current_col <- height_cols[i]
#   
#   for (j in seq_len(nrow(df))) {
#     # last height value
#     prev_value <- NA
#     for (k in (i - 1):1) {  
#       if (!is.na(df[j, height_cols[k]])) {
#         prev_value <- df[j, height_cols[k]]
#         break
#       }
#     }
#     
#     if (!is.na(prev_value) && !is.na(df[j, current_col]) && df[j, current_col] < prev_value) {
#       df[j, current_col] <- NA  
#     }
#   }
# }


```


The dataframe needed to be transformed into a long format dataframe. New variables were also generated: Height, HYY, Year, Age and YAP (years after planting). Some graphs were plotted to check the pattern of the data across ages. 

```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.align='center'}
#head(df); tail(df)
#aggregate(cbind(H95,H96,H97,H98,H99,H00,H01,H02,H03,H04,H05,H10,H15) ~ PROV + REP, FUN = mean, data=df)
#str(df)

# df to long format, variable Height and HYY
df_long <- reshape(df, 
                   varying = list(names(df)[7:19]), # Columns H95 to H15
                   v.names = "Height",              
                   timevar = "HYY",                
                   times = names(df)[7:19],        
                   idvar = "REG",                  
                   direction = "long")

rownames(df_long) <- NULL
head(df_long); tail(df_long)
# df_long$Year <- gsub("H", "", df_long$Year)

#Create Year, Age and YAP variables
df_long$Year <- as.numeric(sub("H", "", df_long$HYY))  
df_long$Year <- ifelse(df_long$Year < 90, df_long$Year + 2000, df_long$Year + 1900)  
df_long$Age <- df_long$Year - 1995 + 6
df_long$YAP <- df_long$Age - 2

# as.factor() required for ASRemlR
# df_long$COL <- as.factor(df_long$COL)
# df_long$ROW <- as.factor(df_long$ROW)
df_long$PROV <- as.factor(df_long$PROV)
df_long$REP <- as.factor(df_long$REP)
df_long$REG <- as.factor(df_long$REG)
df_long$POS <- as.factor(df_long$POS)
#df_long$Age <- as.factor(df_long$Age)

# Data evaluation prior to analysis
table(df$PROV, df$REP)
hist(df_long$Height)
boxplot(Height ~ Age, data = df_long)
boxplot(Height ~ YAP, data = df_long)
meanH_A <- aggregate(Height ~ Age, mean, data = df_long)
varH_A <- aggregate(Height ~ Age, var, data = df_long)
#Mean and var by PROV and Age
meanH_AP <- aggregate(Height ~ PROV + Age, mean, data = df_long)
varH_AP <- aggregate(Height ~ PROV + Age, var, data = df_long)

```

Plotting Age - Height, by provenance
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=11, fig.height=7, fig.align='center'}
labels_data <- meanH_AP %>%
  group_by(PROV) %>%
  filter(Age == max(Age))  # Select the last available Age for each PROV

H_AP_plot <- ggplot(data=meanH_AP, aes(x=Age, y=Height, group = PROV)) +
  geom_line(linewidth=0.7, aes(color=PROV)) + 
  #geom_text(data=labels_data, aes(label=PROV), hjust=-0.1, size=4) + 
  geom_text_repel(data=labels_data, aes(label=PROV), size=4, direction = "y", nudge_x = 2) +  
  ylab("Height (cm)") + xlab("Age (years)") 

H_AP_plot

```

## Data linearisation
The data needed to be transformed (Linearization) according to the previous plots and mean values (different means per PROV and year).  
Model used to transform the data (the intercept and the slope):
$$ln(y_i) = ln(\alpha*) + \frac\beta x_i$$
$$y_i = \alpha + \beta x^*_i + e_i$$
To choose the most appropriate constant "c" used in the standarization of the data, the effect of different constant values were plotted to aid choose the appropriate one. The constants were added to avoid any infinitive value in the data due to the linearization.
```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.width=11, fig.height=7, fig.align='center'}
# Check impact of different c values in linearity
df_imp <- data.frame(Age = 6:26)
df_imp$iAge_0 <- 1 / (df_imp$Age)
df_imp$iAge_3 <- 1 / (df_imp$Age + 3)
df_imp$iAge_5 <- 1 / (df_imp$Age + 5)
df_imp$iAge_7 <- 1 / (df_imp$Age + 7)
df_imp$iAge_10 <- 1 / (df_imp$Age + 10)
df_imp$iAge_20 <- 1 / (df_imp$Age + 20)

df_imp_long <- df_imp %>%
  pivot_longer(cols = starts_with("iAge_"), names_to = "Transformation", values_to = "iAge_Value")

ggplot(df_imp_long, aes(x = Age, y = iAge_Value, color = Transformation)) +
  geom_line(size = 1) +
  labs(
    title = "Impact of Different Transformations on iAge",
    x = "Age (years)",
    y = "Transformed Age (iAge)"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom")

# Linearization (transformation) of the data

# Model
# y = alpha*exp(beta/(x+c))
# ln(y+c) = alpha* + beta*(1/(x+c))

df_long$lnY <- log(df_long$Height+0) # +0, is added to avoid the log of 0 (inf)
df_long$iAge <- 1/(df_long$Age+5)  # (1 / (Age + c)): inverse age transformation 
#log(Age): if growth follows an exponential pattern #scale() to center the mean 0 and scales the var

hist(df_long$Height)
hist(df_long$lnY)
boxplot(Height~Age, data=df_long)
boxplot(lnY~iAge, data=df_long)


mean_lnY_AP <- aggregate(lnY ~ PROV + iAge, FUN = function(x) mean(x, na.rm = TRUE), data = df_long)

lnY_AP_plot <- ggplot(data=mean_lnY_AP, aes(x=iAge, y=lnY, group = PROV)) +
  geom_line(linewidth=0.7, aes(color=PROV)) + 
  ylab("lnY") + xlab("iAge") 

lnY_AP_plot

```

# Random Regression Models

```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.width=11, fig.height=7, fig.align='center'}
# as.factor() required for ASRemlR
# df_long$COL <- as.factor(df_long$COL)
# df_long$ROW <- as.factor(df_long$ROW)
df_long$PROV <- as.factor(df_long$PROV)
df_long$REP <- as.factor(df_long$REP)
df_long$REG <- as.factor(df_long$REG)
df_long$POS <- as.factor(df_long$POS)
df_long$YAP <- as.factor(df_long$YAP)
#df_long$Age <- as.factor(df_long$Age)

table(df_long$PROV, df_long$REP)

```

## General model

The general model used was:
$$y_{ij} = (\alpha + a_i) + (\beta + b_i)x_{ij} + e_{ij}$$
$$y_{ij} = (\alpha + \beta x_{ij}) + (a_i + b_i x_{ij}) + e_{ij}$$
where $\alpha$ (overall intercept) and $\beta$ (overall slope) represent the fixed effect population regression, whereas $a_i$ (random intercept (per PROV)) and $b_i$ (random slope(PER PROV)) represent the random deviations. 


### Model 1. Independent random terms and residuals.

```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.width=11, fig.height=7, fig.align='center'}
rr1 <- asreml(fixed = lnY ~ 1 + REP + iAge,
              random = ~ PROV + PROV:iAge,
              residual = ~ idv(units),
              na.action = na.method(y = "include", x="include"),
              data = df_long)
wald(rr1)
summary(rr1)$varcomp
rr1$loglik # 14064.89
summary(rr1)$aic
summary(rr1)$bic

```

### Model 2. Correlated random effects and independent residuals
The deviations are correlated as:

$$
V\begin{bmatrix}
  a_{i} \\
  b_{i}
\end{bmatrix} = \begin{bmatrix}
  \sigma^2_{a} & \sigma_{ab} \\
  \sigma_{ab} & \sigma^2_{b}
\end{bmatrix}
$$


```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.width=11, fig.height=7, fig.align='center'}
length(levels(df_long$PROV))  ## 18 levels
length(levels(df_long$REG))   ## 1152 levels
asreml.options(maxit=25, gammaPar = FALSE, ai.sing = TRUE, fail="soft", workspace = "4gb")
rr2 <- asreml(fixed = lnY ~ 1 + REP + iAge,
              random = ~str(~PROV + PROV:iAge, ~corgh(2):id(19)),
              residual = ~idv(units),
              na.action = na.method(y = "include", x="include"),
              data = df_long)

wald(rr2)
#summary(rr2)$varcomp
kable(as.data.frame(summary(rr2)$varcomp), format = "html", digits = 6) %>%
  kable_styling(full_width = FALSE) %>%
  scroll_box(width = "100%")
rr2$loglik  #14065.31
summary(rr2)$aic
summary(rr2)$bic
#model comparisson
lrt.asreml(rr1,rr2,boundary = FALSE)

```

### Model 3. Correlated random effects and correlated residuals. The temporal correlation used was first order autoregressive (ar1v).

```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.width=11, fig.height=7, fig.align='center'}
length(levels(df_long$REG))
length(levels(df_long$YAP))
df_long <- df_long[order(df_long$REG, decreasing = FALSE),]
df_long <- df_long[order(df_long$YAP, decreasing = FALSE),]
asreml.options(maxit=25, gammaPar = FALSE, ai.sing = TRUE, fail="soft", workspace = "4gb")

initR <- c(0.5, 0.07) #initial values
rr3 <- asreml(fixed = lnY ~ 1 + REP + iAge,
              random = ~str(~PROV + PROV:iAge, ~corgh(2):id(19)),
              residual = ~ar1v(YAP, init=initR):REG,
              na.action = na.method(y = "include", x="include"),
              data = df_long)

wald(rr3)
#summary(rr3)$varcomp
kable(as.data.frame(summary(rr3)$varcomp), format = "html", digits = 6) %>%
  kable_styling(full_width = FALSE) %>%
  scroll_box(width = "100%")
rr3$loglik # 26621.04
summary(rr3)$aic
summary(rr3)$bic
lrt.asreml(rr2,rr3,boundary = FALSE)
plot(rr3)

# Outliers
res <- resid(rr3)
which.min(res)
df_long[which.min(res),]
df_long[which.max(res),]

```

## Getting outputs (BLUE, BLUP, $\alpha$, $\beta$, $a_i$ and $b_i$) from the model:
```{r, , echo=TRUE, message=FALSE, warning=FALSE, fig.width=11, fig.height=7, fig.align='center'}
#BLUES
BLUES.rr3 <- summary(rr3, coef = TRUE)$coef.fixed

# Getting alpha -> overall intercept
#1/(0+5) 
alpha_int <- predict.asreml(rr3, classify='(Intercept):iAge', levels=list(iAge=0))$pvals  #overall intercept
alpha_int  # 8.466253
# Getting beta -> overall slope
BLUES.rr3
beta_slope <- BLUES.rr3[6,1]  # -45.78893

#BLUPS
BLUP.rr3 <- summary(rr3, coef = TRUE)$coef.random
head(BLUP.rr3); tail(BLUP.rr3)
dim(BLUP.rr3)

BLUP_rr3a <- as.data.frame(BLUP.rr3[1:18,])  #PROV
BLUP_rr3b <- as.data.frame(BLUP.rr3[19:36,])  #PROV:iAge

RRcoef <- data.frame(PROV = rownames(BLUP_rr3a), BLUP_rr3a = BLUP_rr3a$solution, BLUP_rr3b = BLUP_rr3b$solution)

# Getting a and b per PROV
RRcoef$a_int <- alpha_int$predicted.value + RRcoef$BLUP_rr3a
RRcoef$b_slope <- beta_slope + RRcoef$BLUP_rr3b
head(RRcoef); tail(RRcoef)

```

## Predictions and back transforming to the original Ages and Heights
```{r, , echo=TRUE, message=FALSE, warning=FALSE, fig.width=11, fig.height=7, fig.align='center'}
# Predictions for age 40
iAge_40 <- 1/(40+5)
pred_40Age <- predict.asreml(rr3, classify = "PROV:iAge", levels = list(iAge=iAge_40))$pvals
pred_40Age

# Predictions for the measured ages
m_iAges <- unique(df_long$iAge)
m_iAges <- 1/(seq(6,26,1)+5)
pred_iAges <- predict.asreml(rr3, classify = "PROV:iAge", levels = list(iAge=m_iAges))$pvals
pred_Ages <- as.data.frame(pred_iAges)

# Back transforming to original Ages
pred_Ages$Age <- round(1/pred_Ages$iAge - 5.0)
pred_Ages$pred_Height <- exp(pred_Ages$predicted.value)
head(pred_Ages); tail(pred_Ages)
pred_Ages

save.image(file = paste0(config$output_path, config$output_file))

```